---
title: "Statistical Inference On General Social Survey (GSS) Dataset"
author: Muhammad Shoaib Sarwar
output: 
  html_document: 
    fig_height: 4
    highlight: pygments
    theme: spacelab
---

## Setup

### Load Packages

```{r load-packages, message = FALSE}
library(ggplot2)
library(dplyr)
library(statsr)
library(gmodels)
library(scales)
library(sqldf)
```

### Load Data


```{r load-data}
load("gss.Rdata")
```



* * *

## Part 1: Data

```{r}
sqldf("select min(year) as start_year, max(year) as final_year
      from gss")
```

```{r}
dim(gss)
```


The GSS data was collected from 1972 to 2012. There are total 57061 observations and variables are 114. Specific questions (such as national spending priorities, marijuana use, crime and punishment, race relations, quality of life, and confidence in institutions) were asked to the respondents. This implies that there can be missing values since every question is not applicable to each respondent.

As indicated above that a survey was conducted by random sampling, so the results can generalized to the US population. Therefore, the statistical tests performed cannot provide causality relationships between the variables of interest.


* * *

## Part 2: Research Question (RQ)

**RQ:** Suicide opinion disparity exists between genders who have been unemployed in the last ten years from 1972 to 2012.

With the rise of feminism, it would be interesting to analyze whether opinions of females regarding suicide have changed over the years. And, is there similarity among males as well. For this purpose, I compare the proportions of men and women, and draw statistical inferences.

The variables used in this analysis are:

1.  **suicide4:** A categorical variable indicating whether a person is tired of living and ready to die or not.
```{r}
sqldf("select suicide4, count(*) as total_count_of_suicide4_in_gss
      from gss
      group by suicide4")
```

2.  **sex:** A categorical variable indicating the gender of the respondent.
```{r}
sqldf("select sex, count(*) as total_count_of_sex_in_gss
      from gss
      group by sex")
```

3.  **unemp:** A categorical variable referring to unemployment of a person in the last ten years.
```{r}
sqldf("select unemp, count(*) as total_count_of_unemp_in_gss
      from gss
      group by unemp")
```

4.  **year:** From 1972 to 2012.
```{r}
sqldf("select year, count(*) as total_count_of_year_in_gss
      from gss
      group by year")
```

Count aggregations on the above four variables shows that suicide4 and unemp have null values that needs to be fixed. The variable sex have two distinct values i.e. male and females, this is not the case in real life indicating a case of biasness against other genders.

```{r}
gss_suicide <- sqldf("select sex, year, suicide4
                     from gss
                     where (suicide4 is not null and unemp is not null) and unemp = 'Yes'")

dim(gss_suicide)
```


* * *

## Part 3: Exploratory Data Analysis

First of all, checking distribution of suicidal thoughts over the years of people who were unemployed.

```{r}
sqldf("select year, count(suicide4)
      from gss_suicide
      group by year")
```

Year 1983 had most observations with people having suicidal thoughts whereas 1990 is the year with least observations of people who were unemployed and had suicidal thoughts.

Now generating mosaic plot between genders and their suicidal thoughts.

```{r}
mosaicplot(prop.table(table(gss_suicide$sex,gss_suicide$suicide),1),main = 'Gender versus Suicide',
           xlab = 'Gender',ylab = 'Suicidal Thoughts')
```

In the above plot, it can be seen that men have greater proportion of suicidal thoughts as compared to their counter-parts.

```{r}
gss_suicide_prop <- gss_suicide %>%
  group_by(sex,year) %>%
  summarise(prop = sum(suicide4 == "Yes")/n())

ggplot(data=gss_suicide_prop, aes(x=year,y=prop)) + geom_smooth(aes(fill=sex))
```

The loess regression curves for the people who are tired of living and ready to die proportion in men and women is given in the above figure.

The following observations can be made:

*  The peak for men occurred at around 1994 and women at around 1988.

*  The proportion for men is higher as compared to women.

*  The spread of proportion is wider for women.



* * *

## Part 4: Inference

**Null Hypothesis:** The mean difference in proportions of suicidal men and women is zero.

**Alternate Hypothesis:** The mean difference in proportions of suicidal men and women is greater than zero.

One side z-test is performed in-order to test the hypothesis.

First, finding out the pooled proportion as follows:

```{r}
gss_suicide %>%
  summarise(p_pool = sum(suicide4=="Yes")/n(),
            n_1 = sum(sex == "Female"),
            n_2 = sum(sex == "Male"),
            n_1_success = p_pool*n_1,
            n_1_fails = (1-p_pool)*n_1,
            n_2_success = p_pool*n_2,
            n_2_fails = (1-p_pool)*n_2,
            SE = sqrt((p_pool*(1 - p_pool)/n_1) + (p_pool*(1 - p_pool)/n_2)))
```

From the above data, the minimum expected counts criterion of 10 is met for each cell in our 2x2 matrix. The distribution of the sample proportion will be nearly normal, centered at the true population mean.

One side t-test is as follows:

```{r}
inference(y = suicide4, x = sex, data = gss_suicide, statistic = "proportion", type = "ht", null = 0, success="Yes", alternative ="greater", method = "theoretical")

```

From above, it can be seen that the null hypothesis since difference in proportions is 4.0844 standard deviations away from the mean and p-value is less than 0.0001. Hence, it can be easily said that the mean difference in proportions of suicidal men and women is greater than zero.



* * *

## Part 5: Conclusion

In part 4, alternate hypothesis proved to be correct. So, it can be said females in United States have changed their opinions regarding suicide which might be due to the awareness and feminism. On the other hand, proportion of suicidal thoughts in men is proved to be higher.